Explicit Scale Simulation for analysis of RNA-sequencing count data with ALDEx2

### 2025-04-14 go3.Rmd
Penultimate before cleaning

---
title: "Explicit Scale Simulation for analysis of RNA-sequencing  with ALDEx2"
shorttitle: "Scale ALDEx2"
author:
  - name: Greg Gloor
    affiliation: Department of Biochemistry, University of Western Ontario
    email: ggloor@uwo.ca
  - name: Michelle Pistner Nixon
    affiliation: Department of Population Health Sciences, Geisinger, Danville, PA 
  - name: Justin Silverman
    affiliation: 
      College of Information Sciences and Technology and Department of Medicine, Pennsylvania   State University
      Department of Medicine, Pennsylvania State University
      Institute for Computational and Data Science, Pennsylvania State University
      Department of Statistics, Pennsylvania State University
cite-method: citeproc
citeproc: TRUE
bibliography: /Users/ggloor/Library/texmf/bibtex/bib/bibdesk_refs.bib
csl: /Users/ggloor/Documents/0_git/csl_styles/nucleic-acids-research.csl 
output:
    pdf_document:
      keep_tex: true
header-includes:
  - \usepackage{setspace}
package: ALDEx2

@@

\doublespacing 
\singlespacing

High-throughput sequencing (HTS) is a ubiquitous
tool used to explore many biological phenomenon such as gene expression
(single-cell sequencing, RNA-sequencing, meta-transcriptomics), microbial
community composition (16S rRNA gene sequencing, shotgun metagenomics) and
differential enzyme activity (selex, CRISPR killing). HTS proceeds by taking a
sample from the environment, making a library, multiplexing (merging) multiple
libraries together, and then applying a sample of the multiplexed library to a
flow cell. Each of these steps is a compositional sampling step as only a
fixed-size subsample of nucleic acid is carried over to subsequent steps. Thus,
with each sampling step the connection between the actual size of the sampled
DNA pool and the scale (e.g., size, microbial load, or total gene expression) of
the measured biological system is degraded or lost. In the end, the information
contained in the data relates only to relative abundances and  has an arbitrary
scale imposed by the sequencing
process [@lovell:2011;@Quinn:2019aa;@nixon2024scale]. Researchers
can use modified experimental protocols [@Jiang:2011aa;@Loven:2012aa;@Vandeputte:2017aa;@Props:2017aa] or machine learning methods [@Nishijima:2024aa] to uncover the
biological variation in scale. However,  wet-lab protocols only provide information on
the size of the data downstream of the step in the sample preparation protocol where the
intervention was made and  introduce an additional source of variation that must be accounted for [@Loven:2012aa]. The computational methods developed for microbiome analysis have low correlation with the actual scale but are useful [@Nishijima:2024aa]. In short, the disconnect between sample-to-sample variation in sequencing depth and biological variation in scale remains an open
challenge.

The analysis of  HTS data suffers from several known problems that can be
traced, in whole or in part, to  misspecification of scale. The first issue is
poor control of the false discovery rate (FDR) [@Thorsen:2016aa;@Quinn:2018aa;@Nearing:2022aa;@Ge:2021aa;@Li:2022aa], exhibited
as dataset-dependent FDR control and by the disconnect between statistical and
biological significance [@Schurch:2016aa]. In current practice,  these issues are
addressed  by a dual-filtering method, whereby both a low p-value (or
equivalently a low q-value following FDR correction [@storey2003positive]) and a
large difference between groups is used to identify interesting transcripts or
genes for follow-up analysis [@Cui:2003aa;@Schurch:2016aa]. This double-filtering approach is
graphically exemplified by the volcano plot [@Cui:2003aa], but is known to not
appropriately control the  FDR [@Zhang:2009aa;@Ebrahimpoor:2021aa]. The second
issue is poor performance when analyzing data where the mean change between
groups is non-zero [@nixon2024scale]. Such asymmetric data can arise when a gene set is expressed
in one group but not the other, or when one group contains different gene
content from the other. This type of data frequently arises in in-vitro
selection experiments (SELEX), transcriptome analysis, and microbiome analysis
[@Wu2021]. The third issue is that the actual scale of the environment is often
a major confounding variable during analysis [@nixon2024scale;@Nishijima:2024aa]. The final issue is that these problems become more pronounced as more
samples are collected; that is, more information results in a worsening of the
accuracy of the analysis [@Li:2022aa;@nixon2024scale;@Nixon2024B].

The  four problems were recently shown by Nixon et al. [@nixon2024scale]
to be a result of a mismatch between the underlying size or scale of the system
and the assumptions of the normalizations used for the analysis of HTS.
Biological variation in scale often represents an important unmeasured
confounder in HTS analyses [@Lovell:2015]. For example, cells transformed by the
cMyc oncogene have about 3 times the amount of mRNA and about twice the rRNA
content than  non-transformed cells [@Nie:2012aa], and this dramatically skews
transcriptome analysis unless spike-in approaches are used [@Loven:2012aa]. In addition, wild-type and mutant
strains of cell lines, yeast or bacteria  have different growth rates and RNA
contents under different conditions, which  affect our ability to identify truly
differentially abundant genes [@Scott:2010;@Yoshikawa:2011aa;@Lin:2018aa]. As
another example, the total bacterial load of the vaginal microbiome differs by
1-2 orders of magnitude in absolute abundance between the healthy and bacterial
vaginosis states [@Zozaya:2010], and the composition between these states is
dramatically different [@Ravel:2010;@Hummelen:2010]. Thus, a full description of
any of these systems includes both relative change (composition) and absolute
abundance (scale).  Current methods access only the compositional information
yet make implicit assumptions about the scale [@nixon2024scale,@Nixon2024B].

Recently, Nixon et al. [-@nixon2024scale] showed that the challenge of
non-biological variation in sequencing depth be viewed as a problem of
partially-identified models. They showed that \emph{all} normalizations make
some assumption about  scale but these implicit assumptions are often
inappropriate and difficult to interpret. This causes different normalizations
to provide different outputs when applied to the same dataset
[@Bullard:2010;@Dillies:2013;@Thorsen:2016aa;@Weiss:2017aa;@Nearing:2022aa].
Intuitively, normalizations in widespread use  assume that either all samples
have the same scale, e.g. proportions, rarefaction [@Hughes:2005tu], RPKM
[@Mortazavi:2008; @wagner:tpm], etc; or that a subset of features in one sample
can be chosen as a reference to which the others are scaled e.g. the TMM
[@Robinson:2010a], or LVHA [@Wu2021] or the additive log-ratio [@aitchison1982];
or that different sub-parts of each sample maintain a constant scale across
samples e.g. the RLE [@Anders:2010];  or that the geometric mean of the parts is
appropriate e.g the CLR [@Aitchison:1986] and its derivatives.

The original naive ALDEx2 [@fernandes:2013] model unwittingly made a strict
assumption about scale through the CLR normalization [@nixon2024scale]. This assumption was often close enough to the true value to be useful, but was not always the a good estimate and could be outperformed by other normalizations [@Yerke:2024aa].
Nixon et al. [@nixon2024scale] showed that better scale assumptions resulted in
more reproducible data analysis including better control of both false positive
and false negative results. We recently modified ALDEx2  to explicitly model the
scale over a range of reasonable normalization parameters, and showed
significant improvements in performance in microbiome and in-vitro selection
experiments [@Nixon2024B]. Here, we briefly  review these modifications and show how scale uncertainty can greatly improve modeling in transcriptome
and meta-transcriptome datasets to provide more robust and reproducible results.

@@

Formal and expanded descriptions of the concepts that follow are given in
[@nixon2024scale;@Nixon2024B].   To be  concrete, we let \(\mathbf{Y}\) denote
the \emph{measured} \(D \times N\) matrix of sequence counts with elements
\(\mathbf{Y}_{dn}\) indicating the number of measured DNA molecules mapping to
feature \(d\) (e.g., a taxon, transcript or gene) in sample \(n\).  Likewise, we
can denote \(\mathbf{W}\) as the \emph{true}  amount of class \(d\) in the
biological system from which sample \(n\) was obtained. We can think of
\(\mathbf{W}\) as consisting of two parts, the scale \(\mathbf{W}^{\perp}\)
(e.g., totals) and the composition \(\mathbf{W}^{\parallel}\) (i.e.,
proportions). That is, \(\mathbf{W}^{\perp}\) is a \(N\)-vector with elements
\(\mathbf{W}^{\perp}_{n}=\sum_{d}\mathbf{W}_{dn}\) while
\(\mathbf{W}^{\parallel}\) is a \(D \times N\) matrix with elements
\(\mathbf{W}^{\parallel}_{dn}=\mathbf{W}_{dn}/\mathbf{W}^{\perp}_{n}\). Note
that with these definitions \(\mathbf{W}\) can be written as the element-wise
combination of scale and composition:
\(\mathbf{W}_{dn}=\mathbf{W}^{\parallel}_{dn}\mathbf{W}^{\perp}_{n}\), or as the
logarithm \(\log \mathbf{W}_{dn}= \log \mathbf{W}^{\parallel}_{dn} + \log
\mathbf{W}^{\perp}_{n}\).

Many of  the normalizations in widespread use in tools such as DESeq2
[@Love:2014aa], edgeR [@Robinson:2010a], metagenomeSeq [@Paulson:2013aa] ALDEx2
[@fernandes:2014]  can be stated as ratios of the form \(\hat{{\mathbf{W}}}_{dn}
\approx \mathbf{Y}_{dn}/f(\mathbf{Y})\), where the denominator is determined by
some function of the observation. We use the \(\hat{{\mathbf{W}}}\) (\(\ \hat{}\
\)) notation to indicate that the output is an estimate of the true value.  The
technical variation in sequencing depth
(\(\mathbf{Y}^{\perp}_{n}=\sum_{d}\mathbf{Y}_{dn}\)) implies that observed data
\(\mathbf{Y}\) provides us with information about the system composition
\(\mathbf{W}^{\parallel}\) but little to no information in the system scale
\(\mathbf{W}^{\perp}\) (Lovell et al. 2011).

@@

The ALDEx2 R package [@fernandes:2013;@fernandes:2014] is a general purpose
toolbox to  model the uncertainty of HTS data and to use that model to estimate
the underlying LFC (log-fold change) significance.  At a high-level, ALDEx2  has
three connected components to estimate the uncertainty inherent in HTS datasets.
First, the tool accounts for the uncertainty of the sequencing counts using
Dirichlet multinomial sampling to build a probabilistic model of the data; i.e.,
\(\mathbf{\hat{W}}^{\parallel} \approx \mathrm{Dir}(\mathbf{Y})\).  Secondly,
ALDEx2 uses the centred log-ratio transformation to scale the data
[@fernandes:2013]. However, this step was modified recently to account for scale
uncertainty and misspecification [@nixon2024scale,@Nixon2024B] via a scale
model, explained  with more details in [@nixon2024scale;@Nixon2024B] and
summarized in the next paragraph. Finally, a standard null-hypothesis test and a
non-parametric estimate of mean standardized difference are used to report on
the finite sample variation. These sources of uncertainty and variation are
combined via reporting the expected values from a Monte-Carlo simulation
framework. For simplicity, we use the term 'difference' to refer to the absolute
difference between groups, and 'dispersion' to refer to the  within-condition
difference or pooled variance as defined in [@fernandes:2013]. These are
calculated on a \(\log_2\) scale. For more details on ALDEx2 see
[@fernandes:2013;@fernandes:2014;@nixon2024scale;@Nixon2024B].

Scale models can be incorporated into ALDEx2, turning the ALDEx2 model into a
specialized type of statistical model which Nixon et al. [@nixon2024scale] term
a \textit{Scale Simulation Random   Variable} (SSRV). To do this, Nixon et al.
[@nixon2024scale] generalized the concept of normalizations by introducing  the
concept of a \textit{scale model} to account for potential error in the centred
log-ratio normalization step.  They did this by including a model for
\(\mathbf{\hat{W}}^{\perp}_{n}\). The CLR normalization used by ALDEx2 makes the
assumption \(\mathbf{\hat{W}}^{\perp}_{n}=1/G_{n}\), where \(G_n\) is the
geometric mean of sample n, which while being a random variable, is essentially
constant across each Monte-Carlo replicate, but that differs between samples.
With this modification, ALDEx2 can be generalized by considering probability
models for the scale \(\mathbf{\hat{W}}^{\perp}_{n}\) that have mean
\(1/G_{n}\). For example, the following scale model generalizes the CLR:

\[\log \mathbf{\hat{W}}^{\perp}_{n} = -\log G_{n} + \Lambda x_{n} \qquad \Lambda
\sim N(\mu, \gamma^{2})\].

This formulation is quite flexible [@nixon2024scale;@Nixon2024B]. In the simple
or 'default' configuration,   \(\mu = 0\) and  \(\gamma\) is a tunable parameter
drawn from a log-Normal distribution[@nixon2024scale]. Adding scale uncertainty with the \(\gamma\) paramenter  controls only the degree of uncertainty of the CLR assumption for the
\(x_{n}\) binary condition indicator (e.g., \(x_{n}=1\) denotes case and
\(x_{n}=0\) denotes control). In the advanced or 'informed' configuration,
\(\mu\) takes different values for each group and  controls the location of
the LFC assumption; combining \(\mu\) with a \(\gamma\) estimate allows for uncertainty in both the location and the scale. An example of both the default and informed approaches is
given for a microbiome dataset in [@Nixon2024B] showing increased sensitivity and specificity. Here we show that these
approaches also work well in transcriptome and metatranscriptome datasets. These
modifications are  instantiated in ALDEx2 which is  the first software package
designed for SSRV-based inference.

@@

Gierliński et al. [@Gierlinski:2015aa]  generated a  highly replicated yeast
transcriptome dataset to compare gene expression between a wild-type strain and
a snf2 gene knockout, \(\Delta\)snf2. This dataset was  used to test several
RNA-seq tools for their power  to detect the set of differentially abundant
transcripts identified in the full dataset when the data was subset
[@Schurch:2016aa]. In this  study each tool had its own 'gold standard' set of
transcripts with different tools identifying between between 65% to >80% of all
transcripts as being significantly different. Since the majority of transcripts
were significantly different, the authors suggested that it was more appropriate
to apply a  dual cutoff composed of both a Benjamini-Hochberg [@benjamini:1995]
corrected p-value (q-value) plus a difference cutoff  to limit the number of
identified transcripts to a much smaller fraction of the total.  Nixon et al,
@Nixon2024B showed that adding even a small amount of  scale uncertainty with
ALDEx2  dramatically reduced the number of significant transcripts identified,
removing the need for the dual cutoff approach in this dataset and others. Below
we include an intuitive explanation of why and how incorporating scale uncertainty
achieved this outcome using a setting of \(\gamma=0.5\). The approach is in line with
the recommendations of [@Nixon2024B] and  gives results comparable to those
proposed in [@Schurch:2016aa].

```{r yst-res, echo=F, warning=F, message=F,comment=F,}
#ALDEx2
library(ALDEx2) 
load(file="analysis/yst.all.Rda")
load(file="analysis/yst.s.all.Rda") 
load(file="analysis/yst.1.all.Rda")
# DESeq2
load('analysis/res.Rda')

@@

We start with the assumption that not all statistically significant differences
are biologically relevant [@efron2008FDR], and that such a large number of
significantly different transcripts breaks the  necessary assumption for DA/DE
expression that most parts be invariant [@Dillies:2013]. As noted,
transcriptomics commonly uses a dual cutoff approach that is graphically
exemplified by volcano plots [@Cui:2003aa;@Schurch:2016aa]. Using either DESeq2
or ALDEx2, a majority of transcripts are statistically significantly different
between groups  with a  q-value cutoff of \( \le 0.05\); i.e. `r length(sig.des)` (`r round(length(sig.des)/nrow(yst.all), 2)*100`%, DESeq2) or
`r length(sig.ald)` (`r round(length(sig.ald)/nrow(yst.all), 2)*100`%, ALDEx2)
of the `r nrow(yst.all)` transcripts. These values are in line with those
observed by [@Schurch:2016aa]. Such  large numbers of statistically significant
transcripts seems biologically unrealistic. That `r length(setdiff(sig.ald,
sig.des))` transcripts are identified by ALDEx2 and not DESeq2, while DESeq2
identifies `r length(setdiff(sig.des, sig.ald))` transcripts that ALDEx2 does
not, suggests that the choice of normalization plays a role in which results are
returned as significant and that some, if not the majority, are driven by
technical differences in the analysis [@Dillies:2013;@Li:2022aa].

```{r plot1, echo=FALSE, warning=FALSE, fig.dim=c(5,6), fig.cap="Volcano and effect plots for unscaled and scaled transcriptome analysis. ALDEx2 was used to conduct a differential expression (DE) analysis on the yeast transcriptome dataset. The results were plotted to show the relationship between difference and dispersion using effect plots or difference and the q-values using volcano plots. Panels A,C are for the naive analyses, and Panels B,D are for the  default analyses that include scale uncertainty. Each point represents the values for one transcript, with the color indicating if that transcript was significant in the both analyses (red) or in the naive analysis only (orange). Points in grey are not statistically signficantly different under any condition. The horizontal dashed lines represent a \\(\\log_2\\)difference of \\(\\pm 1.4\\)."}

par(mfrow=c(2,2))
# volcano plot(res@listData$log2FoldChange,-1*log10(res@listData$padj + 1e-300),
# col=rgb(0,0,0,0.1), xlab='log2 Difference', ylab='-1 log10(p.adjust)')
# title('D: DESeq2 volcano', adj=0, line= 0.8)
# points(res@listData$log2FoldChange[sig.des],-1*log10(res@listData$padj[sig.des
# ] + 1e-300), col=rgb(1,.66,0,0.3), pch=19, cex=0.5)
# points(res@listData$log2FoldChange[sig.s.ald],-1*log10(res@listData$padj[sig.s
# .ald] + 1e-300), col=rgb(1,0,0,1), pch=19, cex=0.5) abline(v=c(-1.4,1.4),
# lty=2,lwd=2, col='grey')

>>>>

---
title: "Explicit Scale Simulation for analysis of RNA-sequencing count data with ALDEx2"
shorttitle: "Scale ALDEx2"
author:
  - name: Greg Gloor 
    affiliation: Department of Biochemistry, University of Western Ontario 
    email: ggloor@uwo.ca 
  - name: Michelle Pistner Nixon 
    affiliation: Department of Population Health Sciences, Geisinger, Danville, PA 
  - name: Justin Silverman 
    affiliation: College of Information Sciences and Technology and Department of Medicine, Pennsylvania State University, Department of Medicine, Pennsylvania State University, Institute for Computational and Data Science, Pennsylvania State University, Department of Statistics, Pennsylvania State University 
cite-method: citeproc 
citeproc: TRUE 
bibliography: /Users/ggloor/Library/texmf/bibtex/bib/bibdesk_refs.bib 
csl: /Users/ggloor/Documents/0_git/csl_styles/nucleic-acids-research.csl 
output:
    pdf_document:
      keep_tex: true
header-includes:
  - \usepackage{setspace}
  - \usepackage[left]{lineno}

package: ALDEx2

@@

\doublespacing \singlespacing 
\linenumbers

High-throughput sequencing (HTS) is a ubiquitous tool used to explore many biological phenomenon such as gene expression (single-cell sequencing, RNA-sequencing, meta-transcriptomics), microbial community composition (16S rRNA gene sequencing, shotgun metagenomics) and differential enzyme activity (selex, CRISPR killing). HTS proceeds by taking a sample from the environment, making a library, multiplexing (merging) multiple libraries together, and then applying a sample of the multiplexed library to a flow cell. Each of these steps is a compositional sampling step as only a fixed-size subsample of nucleic acid is carried over to subsequent steps. Thus, with each sampling step the connection between the actual number of molecules in the sampled DNA pool and the environmental scale (e.g., total number of molecules, microbial load, or total gene expression) of the measured biological system is degraded or lost. In the end, the information contained in the data relates only to relative abundances and  has an arbitrary scale imposed by the sequencing process [@lovell:2011;@Quinn:2019aa;@nixon2024scale].

The analysis of  HTS data suffers from several known problems that can be traced, in whole or in part, to misspecification of scale in the output data. The first issue is poor control of the false discovery rate (FDR) [@Thorsen:2016aa;@Quinn:2018aa;@Nearing:2022aa;@Ge:2021aa;@Li:2022aa], exhibited as dataset-dependent FDR control which is observed as a disconnect between statistical and biological significance [@Schurch:2016aa;@Li:2022aa]. In current practice, this issue is addressed  by a dual-filtering method, whereby both a low p-value (or equivalently a low q-value following FDR correction [@storey2003positive]) and a large difference between groups is used to identify  transcripts or genes of interest for follow-up analysis [@Cui:2003aa;@Schurch:2016aa]. This double-filtering approach is graphically exemplified by the volcano plot [@Cui:2003aa], but is known to not appropriately control the  FDR [@Zhang:2009aa;@Ebrahimpoor:2021aa]. Since there is no standard way of determining what fold-change cutoff should be used researchers have unlimited degrees of  freedom which is known to lead to unreliable inference [@simmons2011false]. In particular Li et al. [@Li:2022aa] used patient or clinical-derived transcriptome datasets and found that many methods suffer from an extremely high false positive rate. The second issue is poor performance when analyzing data where the mean change between groups is non-zero [@nixon2024scale]. Such asymmetric data can arise when a gene set is expressed in one group but not the other, or when one group contains different gene content from the other. This type of data frequently arises in in-vitro selection experiments (SELEX), transcriptome analysis, and microbiome analysis [@Wu2021]. The third issue is that the actual scale of the environment is often a major confounding variable during analysis [@nixon2024scale;@Nishijima:2024aa]. This has long been known in transcriptome analysis and was a major driver for the development of normalization factors [@Bullard:2010] and the use of molecular spike-ins  to provide a reference set of known counts [@Jiang:2011aa;@Loven:2012aa;@Vandeputte:2017aa;@Props:2017aa] to estimate scale. While often useful, spike-in methods only provide information downstream of the step in the sample preparation protocol where the intervention was made and  introduce an additional source of variation that must be accounted for [@Loven:2012aa]. In the microbiome field, a recent landmark paper showed that biological scale was a major unacknowledged confounder in many human  analyses [@Nishijima:2024aa]. These authors built a machine learning model   to uncover the biological variation in scale and including this information  was useful despite exhibiting only a  modest correlation (estimated to be 0.6)  with the  scale of data external to the training set [@Nishijima:2024aa]. The final issue is that all the above problems become more pronounced as more samples are collected; that is, more information results in optimizing for a precise but inaccurate analysis [@Li:2022aa;@nixon2024scale;@Nixon2024B].

The  four problems were recently shown by Nixon et al. [@nixon2024scale]  to be a result of a mismatch between the underlying size or scale of the system and the assumptions of the normalizations used for the analysis of HTS. Biological variation in scale often represents an important unmeasured confounder in HTS analyses [@Loven:2012aa;@Lovell:2015;@Wu2021;@Nishijima:2024aa]. For example, cells transformed by the cMyc oncogene have about 3 times the amount of mRNA and about twice the rRNA content than  non-transformed cells [@Nie:2012aa], and this dramatically skews transcriptome analysis unless spike-in approaches are used [@Loven:2012aa]. In addition, wild-type and mutant strains of cell lines, yeast or bacteria  have different growth rates and RNA contents under different conditions, which  affect our ability to identify truly differentially abundant genes [@Scott:2010;@Yoshikawa:2011aa;@Lin:2018aa]. As another example, the total bacterial load of the vaginal microbiome differs by 1-2 orders of magnitude in absolute abundance between the healthy and bacterial vaginosis states [@Zozaya:2010], and the cell and RNA composition between these states is dramatically different [@Ravel:2010;@Hummelen:2010]. Thus, a full description of any of these systems includes both relative change (composition) and absolute abundance (scale).  Current methods access only the compositional information yet make implicit assumptions about the scale [@nixon2024scale,@Nixon2024B].

Nixon et al. [-@nixon2024scale] showed that the challenge of non-biological variation in sequencing depth could be explained as a problem of partially-identified models. They showed that \emph{all} normalizations make some assumption about  scale but these implicit assumptions are often inappropriate and difficult to interpret. As a result, different normalizations  provide different outputs when applied to the same dataset [@Bullard:2010;@Dillies:2013;@Thorsen:2016aa;@Weiss:2017aa;@Nearing:2022aa]. Intuitively, normalizations in widespread use  assume that either all samples have the same scale, e.g. proportions, rarefaction [@Hughes:2005tu], RPKM [@Mortazavi:2008; @wagner:tpm], etc; or that a subset of features in one sample can be chosen as a reference to which the others are scaled e.g. the TMM [@Robinson:2010a], the LVHA [@Wu2021]  qPCR [@Vandesompele:2002aa],  the additive log-ratio [@aitchison1982]; or that different sub-parts of each sample maintain a constant scale across samples e.g. the RLE [@Anders:2010];  or that the geometric mean of the parts is appropriate e.g the CLR [@Aitchison:1986] and its derivatives. Notably when the assumption of similar scale across samples is not violated, or violated only weakly,  any method will provide a reasonable analysis. The problem arises  when this assumption is violated, in which case all tools fail without warning.

```{r, out.width="0.6\\linewidth", , echo=FALSE, include=TRUE, fig.align="center", fig.cap=c("Mismatches between estimated scale and true scale lead to poor estimation of high throughput sequencing data. All normalizations used for differential abundance analysis make some strict assumption about the scale of the environment as shown in line 1. In this example all normalizations produce a biased estimate of the environmental scale, but estimate C is the closest to the truth.  Adding uncertainty to normalization C as represented by the distribution in the line 2 leads to less bias and now includes the actual environmental scale in the assumption. As shown in line 3 in some cases it may be useful to adjust the centre of the scale uncertainty estimate if the initial normalizations give very poor estimates of the underlying environment.")}
knitr::include_graphics("./go3_files/figure-latex/scale-explained.pdf")
```

The original scale-naive ALDEx2 [@fernandes:2013] model unwittingly made a strict assumption about scale through the CLR normalization and we found that the CLR was very sensitive to violations of the assumption of scale identity between groups [@Wu2021]. Moreover, when the assumption of identity was not true the CLR used by ALDEx2 could be outperformed by other normalizations in simulation studies [@Yerke:2024aa]. As illustrated graphically in Figure 1, Nixon et al. [@nixon2024scale] showed through simulation that introducing uncertainty in the scale assumptions, and in extreme cases altering the location of the scale assumption, resulted in more reproducible data analysis including better control of both false positive and false negative results. We  modified ALDEx2  to explicitly model uncertainty in scale over a range of reasonable normalization parameters, and showed significant improvements in performance in microbiome and in-vitro selection experiments [@Nixon2024B] and in a vaginal metatranscriptome analysis [@Dos-Santos:2024aa]. Here, we briefly  review these modifications and show how adding scale uncertainty can greatly improve modeling in transcriptome and meta-transcriptome datasets to provide substantially more robust and reproducible results.

@@

Formal and expanded descriptions of the concepts that follow are given in [@nixon2024scale;@Nixon2024B].   To be  concrete, we let \(\mathbf{Y}\) denote the \emph{measured} \(D \times N\) matrix of sequence counts with elements \(\mathbf{Y}_{dn}\) indicating the number of measured DNA molecules mapping to feature \(d\) (e.g., a taxon, transcript or gene) in sample \(n\).  Likewise, we can denote \(\mathbf{W}_{dn}\) as the \emph{true}  amount of class \(d\) in the biological system from which sample \(n\) was obtained. We can think of \(\mathbf{W}\) as consisting of two parts, the scale \(\mathbf{W}^{Tot}\) (e.g., totals) and the composition \(\mathbf{W}^{Comp}\) (i.e., proportions). That is, \(\mathbf{W}^{Tot}\) is a \(N\)-vector with elements \(\mathbf{W}^{Tot}_{n}=\sum_{d}\mathbf{W}_{dn}\) while \(\mathbf{W}^{Comp}\) is a \(D \times N\) matrix with elements \(\mathbf{W}^{Comp}_{dn}=\mathbf{W}_{dn}/\mathbf{W}^{Tot}_{n}\). Note that with these definitions \(\mathbf{W}\) can be written as the element-wise combination of scale and composition: \(\mathbf{W}_{dn}=\mathbf{W}^{Comp}_{dn}\mathbf{W}^{Tot}_{n}\), or as the logarithm \(\log \mathbf{W}_{dn}= \log \mathbf{W}^{Comp}_{dn} + \log \mathbf{W}^{Tot}_{n}\).

Many of  the normalizations used in tools such as DESeq2 [@Love:2014aa], edgeR [@Robinson:2010a], metagenomeSeq [@Paulson:2013aa] ALDEx2 [@fernandes:2014]  can be stated as ratios of the form \(\hat{{\mathbf{W}}}_{dn} \approx \mathbf{Y}_{dn}/f(\mathbf{Y})\), where the denominator is determined by some function of the observation. We use the hat  notation (\(\ \hat{}\ \)) to indicate that the output is an estimate of the true value.  The technical variation in sequencing depth, which is often called "library size" has no relationship with the actual number of molecules in the sampled environment [@lovell:2011]. In other words (\(\mathbf{Y}^{Tot}_{n}=\sum_{d}\mathbf{Y}_{dn}\))  the observed data \(\mathbf{Y}\) provides us with information about the system composition \(\mathbf{W}^{Comp}\) but little to no information in the system scale \(\mathbf{W}^{Tot}\) (Lovell et al. 2011).

@@

The ALDEx2 R package [@fernandes:2013;@fernandes:2014] is a general purpose toolbox to  model the uncertainty of HTS data and to use that model to estimate the significance of the underlying LFC (log-fold change).  At a high-level, ALDEx2  has three connected components to estimate the uncertainty inherent in HTS datasets. First, the tool accounts for the uncertainty of the sequencing counts using Dirichlet multinomial sampling to build a probabilistic model of the data; i.e., \(\mathbf{\hat{W}}^{Comp} \approx \mathrm{Dir}(\mathbf{Y})\).  Secondly, ALDEx2 uses the centred log-ratio transformation to scale the data [@fernandes:2013]. It was this step that was modified  to account for scale uncertainty and misspecification [@nixon2024scale,@Nixon2024B] explained  with more details in [@nixon2024scale;@Nixon2024B] and summarized in the next paragraph. Finally, a standard null-hypothesis test and a non-parametric estimate of mean standardized difference are used to report on the finite sample variation. These sources of uncertainty and variation are combined via reporting the expected values from a Monte-Carlo simulation framework. For simplicity, we use the term 'difference' to refer to the absolute difference between groups, and 'dispersion' to refer to the  within-condition difference or pooled variance as defined in [@fernandes:2013]. These are calculated on a \(\log_2\) scale. For more details on ALDEx2 see [@fernandes:2013;@fernandes:2014;@nixon2024scale;@Nixon2024B].

Scale models were incorporated into ALDEx2, turning the ALDEx2 model into a specialized type of statistical model which Nixon et al. [@nixon2024scale] term a \textit{Scale Simulation Random   Variable} (SSRV). To do this, Nixon et al. [@nixon2024scale] generalized the concept of normalizations by introducing  the concept of a \textit{scale model} to account for potential error in the centred log-ratio normalization step.  They did this by including a model for \(\mathbf{\hat{W}}^{Tot}_{n}\). The CLR normalization used by ALDEx2 makes the assumption \(\mathbf{\hat{W}}^{Tot}_{n}=1/G_{n}\), where \(G_n\) is the geometric mean of the counts (or the corresponding proportions)of each part in sample n, which while being a random variable, is essentially constant across each Monte-Carlo replicate, but that differs between samples. With this modification, ALDEx2 can be generalized by considering probability models for the scale \(\mathbf{\hat{W}}^{Tot}_{n}\) that have mean \(1/G_{n}\). For example, the following scale model generalizes the CLR:

\[\log \mathbf{\hat{W}}^{Tot}_{n} = -\log G_{n} + \Lambda x_{n} \qquad \Lambda \sim N(\mu, \gamma^{2}).\]

This formulation is quite flexible [@nixon2024scale;@Nixon2024B]. In the simple or 'default' configuration,   \(\mu = 0\) and  \(\gamma\) is a tunable parameter drawn from a log-Normal distribution [@nixon2024scale]. Adding scale uncertainty with the \(\gamma\) paramenter (as shown in Figure 1:1) controls only the degree of uncertainty of the CLR assumption for the \(x_{n}\) binary condition indicator (e.g., \(x_{n}=1\) denotes case and \(x_{n}=0\) denotes control). In the advanced or 'informed' configuration, \(\mu\) takes different values for each group and  controls the location of the LFC assumption (as shown in Figure 1:2); combining \(\mu\) with a \(\gamma\) estimate allows for uncertainty in both the location and the scale. In the manuscript where the idea of scale was original derived, Nixon et al. [@nixon2024scale] conducted extensive simulations showing that both the default and informed approaches exhibit   increased sensitivity and specificity [@Nixon2024B]. In this report we show that these approaches also work well to control the FDR in transcriptome and metatranscriptome datasets. We also provide some additional insights into how this is achieved. These modifications are  instantiated in ALDEx2 which is  the first software package designed for SSRV-based inference.

@@

Gierliński et al. [@Gierlinski:2015aa]  generated a  highly replicated yeast transcriptome dataset to compare gene expression between a wild-type strain and a snf2 gene knockout, \(\Delta\)snf2. This dataset of 86 samples (44 and 42 per group) is an example of technical growth replicate experiments common in the literature. The dataset was  used to test several RNA-seq tools for their power  to detect the set of differentially abundant transcripts identified in the full dataset when the data was subset [@Schurch:2016aa]. In this original study each tool had its own 'gold standard' set of transcripts with different tools identifying between between 65% to >80% of all transcripts as being significantly different. Since the majority of transcripts were significantly different, the authors suggested that it was more appropriate to apply a  dual cutoff composed of both a Benjamini-Hochberg [@benjamini:1995] corrected p-value (q-value) plus a difference cutoff  to limit the number of identified transcripts to a much smaller fraction of the total. In other datasets, Nixon et al, @Nixon2024B showed that adding even a small amount of  scale uncertainty with ALDEx2  dramatically reduced the number of significant transcripts identified, removing the need for the dual cutoff approach in this dataset and others.

```{r yst-res, echo=F, warning=F, message=F,comment=F,} 
#ALDEx2 
library(ALDEx2, warn.conflicts=F) 
load(file="analysis/yst.all.Rda") 
load(file="analysis/yst.s.all.Rda") 
load(file="analysis/yst.1.all.Rda") # DESeq2 
load('analysis/res.Rda')

@@

```{r yst-plot, echo=FALSE, warning=FALSE, fig.align="center", out.width="0.8\\linewidth", fig.dim=c(6,7), fig.cap="Volcano and effect plots for unscaled and scaled transcriptome analysis. ALDEx2 was used to conduct a differential expression (DE) analysis on the yeast transcriptome dataset. The results were plotted to show the relationship between difference and dispersion using effect plots or difference and the q-values using volcano plots. Panels A,C are for the naive analyses, and Panels B,D are for the  default analyses that include scale uncertainty. Each point represents the values for one transcript, with the color indicating if that transcript was significant in the both analyses (red) or in the naive analysis only (orange). Points in grey are not statistically signficantly different under any condition. The horizontal dashed lines represent a \\(\\log_2\\)difference of \\(\\pm 1.4\\)."}

load(file="analysis/yst.all.Rda") 
load(file="analysis/yst.s.all.Rda") 
load(file="analysis/yst.1.all.Rda") # DESeq2 
load('analysis/res.Rda')

sig.all <- yst.all$we.eBH < 0.05 
sig.t <- yst.all$we.eBH < 0.05 & abs(yst.all$diff.btw) > 1.4 
sig.s <- yst.s.all$we.eBH < 0.05 
sig.1 <- yst.1.all$we.eBH < 0.05

par(mfrow=c(2,2)) 
# volcano 
# plot(res@listData$log2FoldChange,-1*log10(res@listData$padj + 1e-300), 
#  col=rgb(0,0,0,0.1), xlab='log2 Difference', ylab='-1 log10(p.adjust)')
#   
# title('D: DESeq2 volcano', adj=0, line= 0.8) 
# 
# points(res@listData$log2FoldChange[sig.des],-1*log10(res@listData$padj[sig.des] + 1e-300), 
#   col=rgb(1,.66,0,0.3), pch=19, cex=0.5) 
# points(res@listData$log2FoldChange[sig.s.ald],-1*log10(res@listData$padj[sig.s.ald] + 1e-300),     
#   col=rgb(1,0,0,1), pch=19, cex=0.5) 
#  
# abline(v=c(-1.4,1.4), lty=2,lwd=2, col='grey')
